Simulating High-Dimensional Multivariate Data

### 2021-02-27 020-simulation-algorithm.Rmd
renv+box fully implemented
removed notebooks

```{r ch020-preview, eval=FALSE}
bookdown::preview_chapter('020-simulation-algorithm.Rmd')
```

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$. Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$. To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$. The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.

@@

1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,...,d$.

With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.

@@

There is hope, however, as @GH02 also showed that replicating an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

@@

1. Mapping step
  (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.
  or
  (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Check admissibility and, if needed, compute the nearest correlation matrix.
  (i) Check that $R_X$ is a correlation matrix, a positive [semi]definite matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.
	Otherwise,
  (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
3. NORTA step
  (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
  (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
  (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

@@

$$
\begin{aligned}
\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
\end{aligned}
$$

then there are 1-1 mappings between these sets. We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive. In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PSD matrix with 1's along the diagonal. We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$) and also for checking that each pairwise correlation value is admissible (via the MC estimation of the 2-dimensional Frechet bounds).

@@

The final step implements a high-performance NORTA algorithm. Specifically, step 3i uses a fast multi-core multivariate normal simulator, and steps 3ii, 3iii are parallelized over CPU cores.

@@

*Correlation Conversion*

Given:

* A target Spearman's $\rho$ or Kendall's $\tau$ correlation

Do:

* Determine the Pearson correlation for use in a Gaussian copula such that the resulting random samples using the NORTA algorithm have an estimated Spearman or Kendall correlation that matches the target correlation

From Proposition 22 (Lebrun and Dutfow, 2009), we have the result for Gaussian copula that the copula correlation $r$ is related to Spearman's $\rho$ and Kendall's $\tau$ by the following relations:

$$
r = 2 \sin \left(\frac{\pi}{6} \rho\right) \\
r = \sin \left(\frac{\pi}{2} \tau\right)
$$

* Lebrun, R., & Dutfoy, A. (2009). An innovating analysis of the Nataf transformation from the copula viewpoint. Probabilistic Engineering Mechanics, 24(3), 312-320.

*Pearson Matching*

Given:

* A target Pearson correlation matrix, $R$
* A list of marginal distributions, $F$

Do:

* Determine the input correlation, $\tilde{R}$, such that the resulting random samples using the NORTA algorithm have an estimated Pearson correlation that matches $R$

This method relies on Equations 19, 39, and 49 (Xiao and Zhou, 2019).

* Xiao, Q., & Zhou, S. (2019). Matching a correlation coefficient by a Gaussian copula. Communications in Statistics-Theory and Methods, 48(7), 1728-1747.

@@

$\begin{aligned}
    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
    & X \in S_{+}^{n}
\end{aligned}$

@@

<!-- 
## Other high-performance `bigsimr` algorithms

<mark> ALEX, briefly discuss other algorithms here </mark>
- `cor_bounds()`  
- `cor_fast()`  
- `cor_nearPSD()`

For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \@ref(eq:spearmanRescaled) to further adjust the input correlation matrix. Let the unscaled Spearman correlation coefficients be $\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)$ for two marginal distributions and divide the target correlation by the product in the denominator of Equation \@ref(eq:spearmanRescaled). Let these adjustment factors be denoted as $a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}$ and specifically rescale the target Spearman correlation matrix by

\begin{equation}
(\#eq:convertSpearmanDiscrete)
\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
\end{equation}

In a similar fashion, we rescale Kendall's $\tau$ to adjust the input correlation matrix. The conversion formula is given by

\begin{equation}
(\#eq:convertKendallDiscrete)
\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
\end{equation}

your comment -->

>>>>

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,\ldots,d$. Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$. To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$. The method can be described as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.

@@

1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d$.

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated algorithms --- providing fast scalability using readily-available hardware.

@@

There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

@@

1. Mapping step
    (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.

    or

    (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Check admissibility and, if needed, compute the nearest correlation matrix.
    (i) Check that $R_X$ is a correlation matrix, a positive definite matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.
	Otherwise,
    (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $\tilde{R}_X$, in the Frobenius norm.
3. NORTA step
    (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
    (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
    (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

@@

\begin{align*}
\Omega_P &= \{ R_P : R_P \textrm{ is a correlation matrix for } X \} \\
\Omega_K &= \{ R_K : R_K \textrm{ is a correlation matrix for } X \} \\
\Omega_S &= \{ R_S : R_S \textrm{ is a correlation matrix for } X \} \\
\end{align*}

then there are 1-1 mappings between these sets. We conjecture informally that the sets of admissible $R_S$ and $R_K$ are not highly restrictive. In particular, $R_P$ is approximately $R_S$ for a MVN, suggesting that the admissible set $\Omega_S$ should be flexible as $R_P$ can be any PD matrix with 1's along the diagonal. We provide methods to check whether a target $R_S$ is an element of the *admissible set* $\Omega_S$ (and similarly for $R_K$) and also for checking that each pairwise correlation value is admissible (via the MC estimation of the 2-dimensional Frechet bounds).

@@

The final step implements a high-performance NORTA algorithm. Specifically, step 3i uses a fast multi-core multivariate normal simulator, and steps 3ii, 3iii are distributed over CPU cores.

@@

*Pearson Matching*
Given a desired target correlation $\rho_Y$ between a pair of marginal distributions, the correlation coefficient $\rho_Z$ used in the Gaussian copula must be specified so as to satisfy equation \@ref(eq:pearsonIntegralRelation).

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j
    (\#eq:pearsonIntegralRelation)
\end{equation}

where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables. In general, the double integral in \@ref(eq:pearsonIntegralRelation) is difficult or impossible to solve analytically, so numerical methods are required. Our implementation is based off the work of @xiao2019matching where the marginal transformation is approximated by a weighted sum of Hermite polynomials.

@@

\begin{align*}
    \mathrm{min}\quad & \frac{1}{2} \Vert G - X \Vert^2 \\
    \mathrm{s.t.}\quad & X_{ii} = 1, \quad i = 1, \ldots , n, \\
    & X \in S_{+}^{n}
\end{align*}

@@

### 2021-02-27 index.Rmd
removed notebooks

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package. This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Spearman, or Kendall correlation matrices. It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib", "alex.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package. This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Spearman, or Kendall correlation matrices. It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---
